---
title: "Interactive graphics in R using plotly and leaflet"
output: 
  learnr::tutorial:
  progressive: true
allow_skip: true
df_print: paged
runtime: shiny_prerendered
description: >
  Learn to impress your friends with interactive graphs
---

```{r setup,include=FALSE}
library(leaflet)
library(dplyr)
library(plotly)
library(ggplot2)
library(learnr)


beaven.barley <- read.csv("beanven_barley.csv")


experiment2017<-read.csv("experiment2017.csv")
interview2019<-read.csv("interview2019.csv")

```
  
  
## Overview

In this session we will show you how easy it is to make interactive graphics using R. 

First we will show you how to make any graph from `ggplot2` into an interactive graphic, which you can embed in a website or presentation using the `plotly` library. 

Like this:

```{r introp,echo=FALSE,message=FALSE}
( beaven.barley%>%
  ggplot(aes(x=col, y=gen, colour=yield, row=row))+
    geom_point(size=3)+
      #scale_color_scico(palette = "vik")+
      scale_color_distiller(palette="Spectral")+
        labs(title="Genotype by Column, coloured by yield") ) %>%
          ggplotly()

```



Then we will show how to make interactive maps using the `leaflet` library. 

Like this:

```{r intromap,echo=FALSE,message=FALSE}
leaflet(experiment2017) %>%
  addCircleMarkers(~Longitude, ~Latitude,popup = ~Farmer) %>%
    addProviderTiles('Esri.DeLorme') %>%
      addControl("Location of farms from 2017 experiment",position="topright")
```

Before getting started - we should make sure we install and load the libraries needed for this module. We need the `plotly` library to make our first interactive graph and the `leaflet` library to make our map. We will also use a dataset from the `agridat` library and a colour palette from the `scico` library. Finally, as usual now, we are also using the `ggplot2` and `dplyr` libraries. You should already have them installed, but they still need to be loaded.

```
library(leaflet)
library(plotly)
library(scico)
library(agridat)
library(ggplot2)
library(dplyr)

```



## Interactive Plots Using plotly

Let's say we are working with the `beaven.barley` dataset that we have used during the exercise on modelling. To show the relationship between the genotypes and the yield, while taking into account the strong impact of the column location, we decide to make a scatterplot placing the different genotypes (`gen`) on the y axis, the column location of our experimental design (`col`) on the x axis, and we colour the points according to the yield of barley (`yield`). We needed a strong diverging palette for the colour based on yield and after a few attempts looking at different [options] (https://ggplot2-book.org/scale-colour.html), we find that the `vik` palette from the `scico` package is the one to use.

Here is our plot:

```{r plot1,exercise=TRUE}

beaven.barley%>%
  ggplot(aes(x=col, y=gen, colour=yield))+
    geom_point(size=3)+
      #scale_color_scico(palette = "vik")+
      scale_color_distiller(palette="Spectral")+
        labs(title="Genotype by Column, coloured by yield")

```


This is an interesting graph. We clearly see the effect of the design structure in columns and we also seem to see some yield differences between the genotypes.

Maybe we want to impress our friends, colleagues and supervisors by using this in a presentation. And we could impress them even more by turning it into an interactive plot, in just two simple steps:

* Store the plot as an object 
* Use the function `ggplotly()` on that object 

```{r plot2,exercise=TRUE}

p1 <- beaven.barley %>%
  ggplot(aes(x=col, y=gen, colour=yield))+
    geom_point(size=3)+
      #scale_color_scico(palette = "vik")+
      scale_color_distiller(palette="Spectral")+
        labs(title="Genotype by Column, coloured by yield")

ggplotly(p1)

```


That's it! Although of course remember to install and load the `plotly` library first. 


The key thing to note is that the `+` cannot get you to this point. You have to store the plot as an object and then run the function on that object. 
Alternatively you can wrap your entire plotting code in brackets and pipe from those brackets into `ggplotly()`


```{r plot3, exercise=TRUE}

( beaven.barley%>%
    ggplot(aes(x=col, y=gen, colour=yield))+
      geom_point(size=3)+
        #scale_color_scico(palette = "vik")+
        scale_color_distiller(palette="Spectral")+
          labs(title="Genotype by Column, coloured by yield") ) %>%
            ggplotly()


```

Notice that if you hover over any of the points you see a pop up with the underlying data values.

An additional nice feature you can use here is to add in extra aesthetics which will only appear in the pop-up box. So for example we could add the `row` structure information:

```{r plot4,exercise=TRUE}
( beaven.barley%>%
  ggplot(aes(x=col, y=gen, colour=yield, row=row))+
    geom_point(size=3)+
      #scale_color_scico(palette = "vik")+
      scale_color_distiller(palette="Spectral")+
        labs(title="Genotype by Column, coloured by yield") ) %>%
          ggplotly()

```

Most, but not all, of the things you can customise within `ggplot2` will appear within the interactive plot but sometimes you will come across certain geometries or custom features that do not get converted. And in those cases you tend to get a lot of warning messages, but also an output plot which may be missing some features. 

But, generally, this one additional step works well as a really simple way to liven up a graph for a presentation. Experiment for yourself with any of your own plots!

There are also additional options to customise the interactive features of the plot. You can read more about these <a href="https://plotly-r.com/improving-ggplotly.html" target="_blank">here  </a>  



## Interactive Maps Using leaflet

When we're doing on farm experiments, it is sometimes useful to collect the geographical coordinates of the farms and place these on a map, to see if the geographic landscape can be explain some of the results that we see. 

So as an example, we are going to place the fictive farms of the experiment2017 dataset on a map using the Latitude and Longitude information in the data. Here is a reminder of how the data looks like:

```{r,echo=FALSE}
DT::datatable(experiment2017)

```

I could plot the coordinates using `ggplot2`. But this would be a pretty limited plot, without the context a map would provide.

However the `leaflet` library makes it easy to produce simple, interactive maps.
The number of functions, and options within those functions in `leaflet` can be a little but overwhelming. But it is really simple to make a nice map with just a few simple functions, and mostly sticking to the default options.

To start with you need to set up the data with `leaflet()`. This is quite a lot like the initial `ggplot(data=)` step where we load the data, ready for the next stages.
But by itself, this step just provides a blank canvas.

```{r map0, exercise=TRUE}

leaflet(experiment2017)

```

As you can see - not much happens!

But we can now think about displaying the data. In this case we have farm data that we want to include. A sensible way of plotting points onto a map would be with circles. So we can use a function called `addCircleMarkers()`.

A very important difference to `ggplot2`, is that to add layers within `leaflet` we use the `%>%` pipe and not the `+`.

Within the `addCircleMarkers` function we need to identify the coordinate columns from our data. Longitude first, and then latitude, with both column names preceded by a tilde `~`. And for the later stages of this map to be produced correctly those coordinates must be recorded in decimal degrees. 

So the code looks like this:


```{r map1, exercise=TRUE}
leaflet(experiment2017) %>%
  addCircleMarkers(~Longitude, ~Latitude)
```

Slightly more useful! We can see the position of our farms relative to each other now. But this is what we could have done in `ggplot`. So now what we really want is to see them overlayed on top of a map. We can do this by piping into the `addTiles()` function.

```{r map2, exercise=TRUE}
leaflet(experiment2017) %>%
  addCircleMarkers(~Longitude, ~Latitude) %>%
    addTiles()
```

As you can see - the default is pretty nice here, using the Open Street Map as a base layer. 
We have produced a functional, interactive map in just a few lines. 
As well as the default Open Street Map appearance, `leaflet` makes it easy to bring in base maps from all sorts of third party sources. These are available through the `addProviderTiles()` function, instead of the `addTiles()` function. 

Here we're calling a different base Map ('Esri.DeLorme'), which gives a bit more emphasize to the different features present in the area of our farms:

```{r map3,exercise=TRUE}
leaflet(experiment2017) %>%
  addCircleMarkers(~Longitude, ~Latitude) %>%
    addProviderTiles('Esri.DeLorme')
```

You can see all of the available layers you can use <a href="http://leaflet-extras.github.io/leaflet-providers/preview/" target="_blank">here  </a>  

The scrolling menu on the right hand side of the page shows a preview for these base layers, and just using the short name you can see within that menu within `addProviderTiles()` will let you use this within `leaflet`.

Some of these base layers are not open source, so do require registration and authentication to be able to access, but there are many which can be freely used providing they are appropriately referenced. You will see the references automatically are embedded into the plots, so you don't really even need to think about that step!


It might also be useful to provide a bit more information onto the map.
At the moment if I click on one of my points then nothing happens. But within `addCircleMarkers()` I can add a popup box which shows the content of one of the columns of my data when I click on it. Similar to what we saw with `plotly`.
So I will add the "Farmer" column, to associate the name of the farmer to a farm data point in the map. Again we need a tilde `~` before the name of the column when working in `leaflet`.

```{r mapagain,exercise=TRUE}
leaflet(experiment2017) %>%
  addCircleMarkers(~Longitude, ~Latitude,popup = ~Farmer) %>%
    addProviderTiles('Esri.DeLorme')
```

If you click on any blue circle, now you should see the farmer's name for each farm.

Another useful addition may be to put a title onto this map. The function for this doesn't really have an intuitive name: `addControl()`

```{r anothermap,exercise=TRUE}
leaflet(experiment2017) %>%
  addCircleMarkers(~Longitude, ~Latitude,popup = ~Farmer) %>%
    addProviderTiles('Esri.DeLorme') %>%
      addControl("Location of farms from 2017 experiment")
```

By default this appears in the top left below the zoom controls. I'm not sure I like it there, so instead I will move it to the top right instead.

```{r finalmap,exercise=TRUE}
leaflet(experiment2017) %>%
  addCircleMarkers(~Longitude, ~Latitude,popup = ~Farmer) %>%
    addProviderTiles('Esri.DeLorme') %>%
      addControl("Location of farms from 2017 experiment",position="topright")
```

And I think I am now happy with my map! 

But I haven't even begun to scratch the surface of what leaflet can do. You can find a much more comprehensive tutorial for leaflet here, which will cover all sorts of other ways you can present maps and incorporate your own data into these maps:
<a href="https://rstudio.github.io/leaflet/" target="_blank">https://rstudio.github.io/leaflet/  </a>  

It is worth noting that `leaflet` is a little different to many R packages, since it is a JavaScript package which has been translated into R. And when searching online for `leaflet` you will usually come across people talking about how to write JavaScript code rather than R code. So make sure to include "R" in any search terms when looking for help. 


